---
title: "Race, Age, and Diagnosis Stage in Breast Cancer Mortality"
author: "Echo Tang"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
        self_contained: FALSE
---

Breast cancer is the most common malignant cancer in the world for those assigned female at birth. Like other cancers, breast cancer incidence increases with age with those above 50 years old being more likely to develop breast cancer; however, previous research has also demonstrated that those who are diagnosed at younger ages also have higher reported mortality rates. Previous studies have also shown that race may be a risk factor for breast cancer, as many women of color have higher mortality rates than non-Hispanic white women, and are also associated with getting breast cancer diagnoses at later stages than white women. Given that age and race are two risk factors, there is limited research done on how age at diagnosis can *interact* with stage of diagnosis and race specifically in breast cancer diagnosis, and how that in turn can contribute to and exacerbate these differences in mortality rates. Understanding differences between the clinical progression of breast cancer across these risk factors can contribute to more nuanced care that can potentially bridge inequities in survivability. This study, thereefore, aims to address the question if race affects the age of diagnosis for breast cancer, and if the effects of age and diagnosis stage on vital status differ by race.


```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = TRUE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

```{r load packages, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
```

```{r load-data, echo=FALSE, message=FALSE, results='hide'}
library(SummarizedExperiment)
library(TCGAbiolinks)

clin_query = GDCquery(project = "TCGA-BRCA", data.category = "Clinical", file.type = "xml")
# GDCdownload(clin_query) 
clinic = GDCprepare_clinic(clin_query, clinical.info = "patient")

```


# Results

```{r, message=FALSE, results='hide', warning=FALSE, echo=FALSE}
clinic = as.data.table(clinic) 
skimr::skim(clinic$race_list)
```


```{r, message=FALSE, results='hide', echo=FALSE}
clinic %>% group_by(race_list) %>% summarise(count = n())
109/(nrow(clinic))

```

```{r, message=FALSE, echo=FALSE}
clinic$race_list = as.character(clinic$race_list)
clinic$race_list[clinic$race_list == ""] = NA
clinic$race_list[is.na(clinic$race_list)] = "NOT REPORTED"
clinic$race_list = as.factor(clinic$race_list)

```

```{r, message=FALSE, echo=FALSE}
clinic$race_bin = ifelse(clinic$race_list == "WHITE", "White", ifelse(clinic$race_list == "NOT REPORTED", "Not Reported", "Non-White"))
clinic$race_bin = as.factor(clinic$race_bin)
```

```{r, message=FALSE, results='hide', warning=FALSE, echo=FALSE}

clinic = clinic[, diag_year_imp := fcoalesce(as.double(year_of_initial_pathologic_diagnosis), mean(year_of_initial_pathologic_diagnosis, na.rm = TRUE)), by = .(gender)]
clinic$diag_year_imp = ceiling(clinic$diag_year_imp)

clinic = clinic[, lymph_count := fcoalesce(lymph_node_examined_count, as.integer(mean(lymph_node_examined_count, na.rm = TRUE))), by = .(gender)]

```

```{r, message=FALSE, results='hide', echo=FALSE}
skimr::skim(clinic$diag_year_imp)

```

```{r, message=FALSE, results='hide', echo=FALSE}
sum(is.na(clinic$vital_status))
sum(clinic$vital_status == "Dead")

```

```{r, message=FALSE, results='hide', echo=FALSE}
clinic$after_diagnosis = ifelse(is.na(clinic$days_to_death), 2022 - clinic$diag_year_imp, ceiling(clinic$days_to_death/365 ))

summary(clinic$after_diagnosis)

```

```{r, message=FALSE, results='hide', echo=FALSE}
sum(is.na(clinic$stage_event_pathologic_stage))

```

```{r, message=FALSE, results='hide', echo=FALSE}
dim(clinic)
str(clinic)
head(clinic)
tail(clinic)

```

After data wrangling and cleaning, the resulting dataset consists of 1062 observations for 116 variables, of which include newly imputed categorical variables and the years survived after diagnosis variable. 

```{r, message=FALSE, results='hide', echo=FALSE}
summary(clinic$stage_event_pathologic_stage)
table(clinic$stage_event_pathologic_stage)

summary(clinic$race_list)
table(clinic$race_list)

summary(clinic$age_at_initial_pathologic_diagnosis)

summary(clinic$diag_year_imp)

summary(clinic$after_diagnosis)
```

```{r, message=FALSE, echo=FALSE}
clinic$stage_event_pathologic_stage = ifelse(clinic$stage_event_pathologic_stage == "Stage I" | 
                          clinic$stage_event_pathologic_stage == "Stage IA" | 
                          clinic$stage_event_pathologic_stage == "Stage IB", "Stage I", 
      ifelse(clinic$stage_event_pathologic_stage == "Stage II" | 
                          clinic$stage_event_pathologic_stage == "Stage IIA" | 
                          clinic$stage_event_pathologic_stage == "Stage IIB", "Stage II", 
      ifelse(clinic$stage_event_pathologic_stage == "Stage III" | 
                          clinic$stage_event_pathologic_stage == "Stage IIIA" | 
                          clinic$stage_event_pathologic_stage == "Stage IIIB" | 
                          clinic$stage_event_pathologic_stage == "Stage IIIC", "Stage III", 
      ifelse(clinic$stage_event_pathologic_stage == "Stage IV", "Stage IV", "Stage X"))))
```

```{r, echo=FALSE}
stage_stats = clinic %>% 
  group_by(stage_event_pathologic_stage) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

stage_stats = stage_stats %>% rename("count" = "Number of Patients", "stage_event_pathologic_stage" = "Stage of Diagnosis")
  
```

## Exploratory Data Analysis

```{r, message=FALSE, echo=FALSE}

age_prelim = clinic %>%
  group_by(race_bin) %>%
  do(p=plot_ly(., x = ~vital_status, y = ~age_at_initial_pathologic_diagnosis, color = ~race_bin, type = "box", boxmean = TRUE)) %>% 
  subplot(nrows = 1, shareY = TRUE) %>% layout(title = "Age at Diagnosis Stratified by Vital Status Across Racial Groups", yaxis = list(title = "Age at Diagnosis in Years", hoverformat = '.2f'))

age_prelim

```

There are more non-white patients who have passed away with younger diagnosis ages than white and non-white patients; however, the mean and median age of diagnosis among white and non-white patients who have passed away are still quite similar. Regardless of vital status, patients with issing race information have higher ages of diagnosis with the mean age of diagnosis being much higher than the median for those who have passed away. 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
vit_age = clinic %>% 
  ggplot(mapping = aes(x = age_at_initial_pathologic_diagnosis, y = after_diagnosis, color = vital_status)) + 
  geom_jitter(aes(size = after_diagnosis, alpha = 0.5, text = paste(paste("Vital status: ", vital_status), "\n", paste("Years Survived after Diagnosis: ", after_diagnosis)))) + 
  geom_smooth(method = "lm") + 
  facet_wrap(~stage_event_pathologic_stage) + 
  xlab("Age in Years at Diagnosis") + 
  ylab("Years Survived After Diagnosis") + 
  ggtitle("Years Survived After Cancer Diagnosis by Age and Vital Status") + guides(alpha = FALSE) 

ggplotly(vit_age, tooltip = "text") %>% layout(legend = list(title = list(text = "Vital Status")))

```

For all stages, those who have passed away have less years survived after diagnosis across all ages compared to those who are still alive. However, some relationships between years survived after diagnosis and age at diagnosis differ across stage; although there seems to be a slight negative relationship between years survived after diagnosis and diagnosis age for stages I, II, III, and X across vital status, there is a slightly positive relationship among those diagnosed at Stage IV across vital status. 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
age_race_stage = clinic %>% 
  ggplot(mapping = aes(x = age_at_initial_pathologic_diagnosis, y = after_diagnosis, color = race_bin)) + 
  geom_jitter(aes(text = paste("Vital status: ", vital_status))) + 
  geom_smooth(method = "lm") + 
  facet_wrap(~stage_event_pathologic_stage) + 
  xlab("Age in Years at Diagnosis") + 
  ylab("Years Survived After Diagnosis") + 
  ggtitle("Years Survived After Cancer Diagnosis by Age, Race, and Stage") 

ggplotly(age_race_stage, tooltip = "text") %>% layout(legend = list(title = list(text = "Race")))

```


There seems to be very slightly negative relationship between years survived after diagnosis and age in stages I, II, and III across all racial groups. There is, however, a slight positive relationship between years survived after diagnosis and age for white and missing race patients in Stage IV and non-white patients in Stage X. This suggests that the relationship between years survived and age does differ to an extent by stage.  

These visual differences suggest that race of a patient and stage of diagnosis may have interaction effects; however, a bigger sample size may be needed to confirm these relationships.


```{r, message=FALSE, warning=FALSE, echo=FALSE}
year_clinic = clinic %>% group_by(race_bin, vital_status) %>% mutate(mean_years = round(mean(after_diagnosis), 2))

summary = year_clinic %>% 
  ggplot() + 
  stat_summary(mapping = aes(x = race_bin, y = after_diagnosis, color = race_bin, text = paste(paste("Vital Status: ", vital_status, "\n"), paste("Mean years survived after diagnosis: ", mean_years))), 
               fun.min = min, fun.max = max, fun = mean, na.rm = TRUE) + 
  xlab("Race") + 
  ylab("Years Survived After Diagnosis") + 
  ggtitle("Summary of Years Survived After Diagnosis by Race") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), plot.margin = margin(10, 10, 10, 100))

ggplotly(summary, tooltip = "text") %>% layout(legend = list(title = list(text = "Race")))

```

The summary graph suggests that white patients who have survived have on average live longer than non-white patients who have survived; on the other hand, the average years survived after diagnosis for those have passed away do not differ much between non-white and white patients. However, the mean years survived after diagnosis for the missing race patients who have survived is slightly higher than non-white and white-patients, and the mean years survived after diagnosis for those who have passed away are lower than non-white and white patients. Furthermore, the range of years survived after diagnosis for both white and non-white patients are much larger than those who have missing race information. Given the relationships between years survived after diagnosis and age of diagnosis for stage IV and X for non-reported and white patients, it is likely that missing race patients are the most similar to white patients; however, because of similarities in age of diagnosis across vital status and distributions in diagnosis stage for non-white and white patients in this dataset, this cannot be confirmed.  


# Conclusion
From this analysis, it was revealed that there were no significant differences between the distribution of diagnosis age or stage of diagnosis among white and non-white breast cancer patients. However, it was found that across vital status, the relationship between years survived after diagnosis for those who were diagnosed in Stage IV differs from other stages of diagnosis where there was a slightly positive relationship compared to other stages that have seen slightly negative relationships. Similarly, white and missing race patients saw a slightly positive relationship between years survived after diagnosis and age of diagnosis in Stage IV compared to other stages, where patients of all racial groups saw a slightly negative relationship. However, due to the limited number of patients in both Stage IV compared to the other stages, future analysis should be done with a larger sample size. 

# References
Azamjah, N., Soltan-Zadeh, Y., & Zayeri, F. (2019). Global trend of breast cancer mortality
rate: A 25-year study. Asian Pacific Journal of Cancer Prevention, 20(7), 2015–2020.

McGuire, A., Brown, J., Malone, C., McLaughlin, R., & Kerin, M. (2015). Effects of age on the detection and management of  breast cancer. Cancers, 7(2), 908–929. 

National Institute of Health. (n.d.). The Cancer Genome Atlas Program. National Cancer
Institute. Retrieved October 18, 2022, from
https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga

Ooi, S. L., Martinez, M. E., & Li, C. I. (2010). Disparities in breast cancer characteristics and outcomes by Race/Ethnicity. Breast Cancer Research and Treatment, 127(3), 729–738. 

Wang, L. (2017). Early Diagnosis of Breast Cancer. Biosensors for Cancer Biomarkers, 17(7),
1572. 

